Update of the ozone dry deposition in the OsloCTM3 v1.0

### 2018-10-10 DryDep.tex
Add drawing of mapping of different landuse categories. Copied all text about Ra and almost all about Rb from manual. Fixed bug in make. TODO: Complete bibliography.

Tropospheric ozone is mainly produced in situ in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) or volatile organic substances (VOCs - also known as hydrocarbons) in the presents of nitrogen oxides (\chem{NO_x}). A typical reaction mechanism for \chem{CO} is sketched below. In a sequence of rapid reactions a peroxyl radical \chem{HO_2^\bullet} is formed through an initial reaction of \chem{CO} with a hydroxyl radical \chem{^\bullet OH}
  ^\bullet\chem{OH} + \chem{CO} \rightarrow {^\bullet\chem{HOCO}}\\
  ^\bullet\chem{HOCO} + \chem{O_2} \rightarrow \chem{HO_2^\bullet} + \chem{CO_2}.
Via a reaction between \chem{HO_2^\bullet} and \chem{NO}, \chem{NO_2} is formed which is then photolyzed. The resulting atomic oxygen reacts then with \chem{O_2} (and any other available co-reactant $M$) to form an ozone molecule
  \chem{HO_2^\bullet} + \chem{NO} \rightarrow {^\bullet\chem{OH}} + \chem{NO_2}\\
  \chem{NO_2} + h\nu \rightarrow \chem{NO} + \chem{O(^3{P})}\\
  ^\bullet\chem{O(^3P)} + \chem{O_2} + M \rightarrow \chem{O_3} + M.

@@

Similar cycles involving volatile organic compounds (VOCs) exist \citep{ACP:Monks2015}. Another source of tropospheric ozone is downward transport from the stratosphere via stratosphere-troposphere exchange (STE) \citep{WMO2014}. Based on observations, STE might only amount to roughly 10\,\unit{\%} ($550 \pm 140$\,\unit{Tg a^{-1}}) of the total global ozone budget in the troposphere, while ozone from chemical production is estimated to be 5000\,\unit{Tg a^{-1}} \citep{ACP:Monks2015}. Ozone is removed from the atmosphere by photochemical reactions or scavenging processes. Major sinks are photolysis followed by a reaction with water vapor to from \chem{OH}
  \chem{O_3} + h\nu \rightarrow \chem{O(^1D)} + \chem{O_2})\\
  \chem{O(^1D)} + \chem{H_2O} \rightarrow 2\chem{OH},
reactions with \chem{HO_2} \citep{ACP:Seinfeld2006}
  {\bf TODO: Reaction},
de-nitration reactions via
  \chem{NO_x} + \chem{O_3} \rightarrow \chem{NO_{x+1}} + \chem{O_2},

@@

@@

wherein aerodynamic $R_a$, quasi laminar $R^i_b$, and canopy resistances $R^i_c$ are key components \citep{AE:Wesely1989}. For all gases $R_a$ is the same, while $R^i_b$ and $R^i_c$ vary from gas to gas and also depend on landuse types (e.g., ice/snow, water, urban, desert, agricultural land, deciduous forest, coniferous forest etc.). Originally, \citet{AE:Wesely1989} used fixed seasonal average dry deposition resistances for each landuse type. For all three types of resistances in this Wesely-type parameterization, more process-oriented formulations have been developed and validated over the years. \citet{ACP:Luhar2017} have validated ozone dry deposition to the ocean with respect to three different formulations of surface resistances. An update on the ozone surface resistance over snow and ice covered surfaces has been provided from combined model and observation studies \citep[e.g., $v^\chem{O_3}_\text{ice/snow} = 1/10000\,\unit{cm s^{-1}}$,][]{ACP:Helmig2007}. Canopy conductance is formulated at the single-leave level (stomatal conductance) for various plant function types (PFT) as well as single species based on empirical studies \citep{PTRS:Jarvis1976, BallBerry1987, ACP:Simpson2012, ICP:MappingManual2017}. But progress has also been made on process oriented modeling of stomatal conductance \citep{PP:Buckley2017}. The variety of differing formulations and choices of parameters leads to a wide spread in model intercomparisons \citep{ACP:Hardacre2015} and about 20\,\unit{\%} uncertainty on the resulting total dry deposition \citep{ACP:Monks2015}.\\

@@

\section{Model description}
\label{sec:model_des}
The OsloCTM3 is a three dimensional global chemistry transort model (CTM). The key components of the OsloCTM3 have been described and evaluated by \citet{GMD:Sovde2012}. A detailed account of the capabilities of the OsloCTM3 in simulating anthropogenic aerosol forcing in the past and recent past using the Community Emission Data System (CEDS) historical emission inventory \citep{GMD:Hoesly2018} is given by \citet{GMD:Lund2018}. \emph{The OsloCTM3 can also be coupled to MEGAN \citep{ACP:Guenther2006}. A publication focussing on this is planed.}
Technical: Partly parallelized (low number of effective cores 16, huge memory consumption if run on T159 resolution, code has been ported to fortran90 keeping the dynamical core UCI-CTM mainly intact.)
historical anthropogenic and biomass burning emissions \citep{ACP:Lamarque2010}, meteorological driver \href{https://www.ecmwf.int/en/forecasts/documentation-and-support/evolution-ifs/cycle-38r1-summary-changes}{ECMWF - OpenIFS version cy38r1nc4}, Community Emission Data System (CEDS) historical emission inventory \citep{GMD:Hoesly2018}

@@

We follow the Convention on Long-Range Transboundary Air Pollution (CLRTAP) implemented in the European Monitoring and Evaluation Programme (EMEP) MSC-W model \citep{ACP:Simpson2012,ICP:MappingManual2017}. It is a more physical approach compared to \citet{AE:Wesely1989, JGR:Hough1991}. The scheme is used for the gaseous species \chem{O_3}, \chem{H_2O_2}, \chem{NO_2}, \chem{PAN}, \chem{SO_2}, \chem{NH_3}, \chem{HCHO}, and \chem{CH_3CHO}. Since \chem{CO} has a very small uptake and is not included in the EMEP scheme, the old OsloCTM2 parameterization is kept. In adition to the gaseous species, some of the aerosol deposition rates have also been modified, namely black carbon (BC) and organic carbon (OC), sulphuric (\chem{SO_4}, \chem{MSA}), and secondary organic aerosols (SOA).

>>>>

Tropospheric ozone is mainly produced in situ in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) or volatile organic substances (VOCs - also known as hydrocarbons) in the presents of nitrogen oxides (\chem{NO_x}). A typical reaction mechanism for \chem{CO} is sketched below. In a sequence of rapid reactions a peroxyl radical \chem{HO_2^\bullet} is formed through an initial reaction of \chem{CO} with a hydroxyl radical \chem{^\bullet OH}.
%\begin{reaction}
%  ^\bullet\chem{OH} + \chem{CO} \rightarrow {^\bullet\chem{HOCO}}\\
%  ^\bullet\chem{HOCO} + \chem{O_2} \rightarrow \chem{HO_2^\bullet} + \chem{CO_2}.
%\end{reaction}
Via a reaction between \chem{HO_2^\bullet} and \chem{NO}, \chem{NO_2} is formed which is then photolyzed. The resulting atomic oxygen reacts then with \chem{O_2} (and any other available co-reactant $M$) to form an ozone molecule.
%\begin{reaction}
%  \chem{HO_2^\bullet} + \chem{NO} \rightarrow {^\bullet\chem{OH}} + \chem{NO_2}\\
%  \chem{NO_2} + h\nu \rightarrow \chem{NO} + \chem{O(^3{P})}\\
%^\bullet\chem{O(^3P)} + \chem{O_2} + M \rightarrow \chem{O_3} + M.
%\end{reaction}

@@

Similar cycles involving volatile organic compounds (VOCs) exist \citep{ACP:Monks2015}. Another source of tropospheric ozone is downward transport from the stratosphere via stratosphere-troposphere exchange (STE) \citep{WMO2014}. Based on observations, STE might only amount to roughly 10\,\unit{\%} ($550 \pm 140$\,\unit{Tg a^{-1}}) of the total global ozone budget in the troposphere, while ozone from chemical production is estimated to be 5000\,\unit{Tg a^{-1}} \citep{ACP:Monks2015}. Ozone is removed from the atmosphere by photochemical reactions or scavenging processes. Major sinks are photolysis followed by a reaction with water vapor to from \chem{OH},
%\begin{reaction}
%  \chem{O_3} + h\nu \rightarrow \chem{O(^1D)} + \chem{O_2})\\
%  \chem{O(^1D)} + \chem{H_2O} \rightarrow 2\chem{OH},
%\end{reaction}
reactions with \chem{HO_2} \citep{ACP:Seinfeld2006},
%\begin{reaction}
%  {\bf TODO: Reaction},
%\end{reaction}
de-nitration reactions,
%\begin{reaction}
%  \chem{NO_x} + \chem{O_3} \rightarrow \chem{NO_{x+1}} + \chem{O_2},
%\end{reaction}

@@

  \label{eq:drydep_velo}

@@

wherein aerodynamic $R_a$, quasi-laminar layer $R^i_b$, and canopy resistances $R^i_c$ are key components \citep{AE:Wesely1989}. For all gases $R_a$ is the same, while $R^i_b$ and $R^i_c$ vary from gas to gas and also depend on landuse types (e.g., ice/snow, water, urban, desert, agricultural land, deciduous forest, coniferous forest etc.). Originally, \citet{AE:Wesely1989} used fixed seasonal average dry deposition resistances for each landuse type. For all three types of resistances in this Wesely-type parameterization, more process-oriented formulations have been developed and validated over the years. \citet{ACP:Luhar2017} have validated ozone dry deposition to the ocean with respect to three different formulations of surface resistances. An update on the ozone surface resistance over snow and ice covered surfaces has been provided from combined model and observation studies \citep[e.g., $v^\chem{O_3}_\text{ice/snow} = 1/10000\,\unit{cm s^{-1}}$,][]{ACP:Helmig2007}. Canopy conductance is formulated at the single-leave level (stomatal conductance) for various plant function types (PFT) as well as single species based on empirical studies \citep{PTRS:Jarvis1976, BallBerry1987, ACP:Simpson2012, ICP:MappingManual2017}. But progress has also been made on process oriented modeling of stomatal conductance \citep{PP:Buckley2017}. The variety of differing formulations and choices of parameters leads to a wide spread in model intercomparisons \citep{ACP:Hardacre2015} and about 20\,\unit{\%} uncertainty on the resulting total dry deposition \citep{ACP:Monks2015}.\\

@@

\section{Model description}
\label{sec:model_des}
The OsloCTM3 is a three dimensional global chemistry transport model (CTM). The key components of the OsloCTM3 have been described and evaluated by \citet{GMD:Sovde2012}. A detailed account of the capabilities of the OsloCTM3 in simulating anthropogenic aerosol forcing in the past and recent past using the Community Emission Data System (CEDS) historical emission inventory \citep{GMD:Hoesly2018} is given by \citet{GMD:Lund2018}. \emph{The OsloCTM3 can also be coupled to MEGAN \citep{ACP:Guenther2006}. A publication focussing on this is planed.}
Technical: Partly parallelized (low number of effective cores 16, huge memory consumption if run on T159 resolution, code has been ported to fortran90 keeping the dynamical core UCI-CTM mainly intact.)
historical anthropogenic and biomass burning emissions \citep{ACP:Lamarque2010}, meteorological driver \href{https://www.ecmwf.int/en/forecasts/documentation-and-support/evolution-ifs/cycle-38r1-summary-changes}{ECMWF - OpenIFS version cy38r1nc4}, Community Emission Data System (CEDS) historical emission inventory \citep{GMD:Hoesly2018}

@@

We follow the Convention on Long-Range Transboundary Air Pollution (CLRTAP) implemented in the European Monitoring and Evaluation Programme (EMEP) MSC-W model \citep{ACP:Simpson2012,ICP:MappingManual2017}. It is a more physical approach compared to the previous used Wesely scheme. The EMEP scheme is used for the gaseous species \chem{O_3}, \chem{H_2O_2}, \chem{NO_2}, \chem{PAN}, \chem{SO_2}, \chem{NH_3}, \chem{HCHO}, and \chem{CH_3CHO}. Since \chem{CO} has a very small uptake and is not included in the EMEP scheme, the old OsloCTM2 \citep{OsloCTM2} parameterization is kept. In adition to the gaseous species, some of the aerosol deposition rates have also been modified, namely black carbon (BC) and organic carbon (OC), sulphuric (\chem{SO_4}, \chem{MSA}), and secondary organic aerosols (SOA). In Eq.~(\ref{eq:drydep_velo}), the dry deposition computation is splitted into three different parts which will be addressed in the following.

\subsubsection{Aerodynamic resistance}
\label{subsubsec:Ra}
The aerodynamical resistance $R_a$ as described in \citet{Simpson2003,ACP:Simpson2012}
\begin{equation}
  R_a = \frac{1}{\kappa u_*}\left[{\ln{\left(\frac{z-d}{z_0}\right)}-\Psi_m\left(\frac{z-d}{z_0}\right)-\Psi_m\left(\frac{z_0}{L}\right)}\right],
\end{equation}
with $\kappa$ the K\'{a}rm\'{a}n constant, $u_*$ the friction velocity, $\Psi_m$ the integrated stability equations for momentum, $d$ a constant (typically 0.7\,\unit{m}), and $L$ the Obukhov length. For certain values of $z$, $z_0$, and $L$, this may result in unphysical (negative) values for $R_a$. For this reason, we diverge from the EMEP scheme at this point and follow the method of \citet{Monteith1973}.
The sensible heat flux $\Phi_{Q_\text{sensible}}$ can be written as
\begin{equation}
\Phi_{Q_\text{sensible}} = \rho_\text{air} c_P \cdot \frac{\partial_z T}{\partial_z R_a},
\end{equation}
wherein $\rho_\text{air}$ is the air density, $c_P$ the specific heat at constant pressure $P$, $\partial_z T$ is the difference between the surface temperature $T_0$ and the temperature at reference height $T_z$, and $\partial_z R_a$ the diffusion resistance to sensible heat between surface and the reference height. From the eddy diffusion theorem, a similar formulation can be derived
\begin{equation}
  \Phi_{Q_\text{sensible}} = \rho_\text{air} c_P \cdot \frac{\partial_z T}{\partial_z u} \cdot u_*^2.
\end{equation}
From this and assuming $\partial_z R_a \rightarrow R_a $ and $\partial_z u \rightarrow u_z $ for a finite $z$, we find 
\begin{equation}
  R_a = \frac{u_z}{u_*^2}.
\end{equation}
As reference height $z$ we use the height at midlevel of the lowermost model level (roughly 8\,\unit{m}) and $u_z$ and $u_*$ are available from the meteorological input fields.

\subsubsection{Quasi-laminar layer resistance}
\label{subsubsec:Rb}
$R_b^i$ is species specific, as already mentioned, and its definition differs over land and ocean. In case a gridbox contains both land and ocean, a weighted mean is calculated using the land and ocean conductances, respectively.\\

Over land we use
\begin{equation}
  R_b^i = \frac{2}{\kappa u_*} \cdot \left(\frac{\text{Sc}_i}{\text{Pr}}\right)^{\frac{2}{3}},
\end{equation}
wherein $\text{Pr}$ is the Prandtl number (typically 0.72 for air and other gases) and $\text{Sc}_i$ is the Schmidt number for a gas $i$. Assuming a constant kinetic viscosity of air $\nu_\text{air}$, $\nu_\text{air}$ can be calculated from the Schmidt number for water $\text{Sc}_\chem{H_2O}=0.6$ and a molecular diffusivity of $D_\chem{H_2O}=0.21\cdot10^{-4}\,\unit{m^2s^{-1}}$
\begin{equation}
  \text{Sc}_i = \frac{\nu}{D_i} = \text{Sc}_\chem{H_2O}\cdot\frac{D_\chem{H_2O}}{D_i},
\end{equation}
with $D_i$ the molecular diffusivity for a gas $i$. The ration $D_\chem{H_2O}/D_i$ is given in Table~\ref{tab:diffusivity}.\\
%
\begin{table}[t]
  \caption{Molecular diffusivity for a gas $i$ given as ration $D_\chem{H_2O}/D_i$.}
  \begin{tabular}{lc}
    \tophline
     & $D_\chem{H_2O}/D_i$\\
    \middlehline
    \chem{O_3} & 1.6\\
    \chem{SO_2} & 1.9\\
    \chem{NO_2} & 1.6\\
    \chem{H_2O_2} & 1.4\\
    \chem{HCHO} & 1.3\\
    \chem{NO} & 1.3\\
    \chem{CH_3CHO} & 1.6\\
    \bottomhline
  \end{tabular}
  %\belowtable{} % Table Footnotes
  \label{tab:diffusivity}
\end{table}

Over ocean we use
\begin{equation}
  R_b^i = \frac{1}{\kappa u_*}\cdot\ln\left({\frac{z_0}{D_i}\cdot \kappa u_*}\right)
\end{equation}
with a minimum auf 10\,\unit{ms^{-1}} and a maximum of 1000\,\unit{ms^{-1}}. The computation of roughneth length $z_0$ over ocean is devided into a \emph{calm} and a \emph{rough} ocean case, where windspeeds above 3\,\unit{ms^{-1}} are considered as boundary between the cases. Calm sea follows \citet{Hinze1975,Garrat1992}:
\begin{equation}
  z_0^\text{calm water} = \text{min}\left(2\cdot10^{-3}, 0.135 \frac{\nu}{u_*}\right),
\end{equation}
where $\nu$ is the kinematic viscosity of air, computed from surface pressure $P_0$, 2\,\unit{m} temperature $T\text{2M}$, and universal gas constant for air $R_\text{air}$
\begin{equation}
  \nu = \frac{6.2\cdot 10^{-8} \cdot T_\text{2M}}{\frac{P_0}{T_\text{2M}\cdot R_\text{air}}}.
\end{equation}
\subsubsection{Surface resistance}
\label{subsubsec:Rc}
%\begin{figure}[t]
%  \centering
%  \includegraphics[width=8.3cm]{pictures/pft_categories.pdf}\\
%  \includegraphics[width=8.3cm]{pictures/pft_categories_zonal.pdf}
%\caption{Plant functional types (PFT). (a) CML2 dynamic landuse 0.5x0.5\,\unit{^\circ}}
%\label{fig:pft_landuse}
%\end{figure}
Mapping of pft landuse types from either ISLSCP2 MODIS or CML2 categories to 9 classical biosphere types used in the EMEP scheme. EMEP categories: Forests, Mediterranean scrub, Crops, Moorland (savanna++), Grassland, Wetlands, Tundra, Desert, Water, Urban
Weighting of conductance with landuse fraction and sum. Inverse of this is gridbox average resistance.
